Potential and Pitfalls of Multi-Armed Bandits for Decentralized Spatial Reuse in WLANs

### 2018-11-16 LaTeX files/potential_pitfalls_mabs_spatial_reuse.tex
Version accepted (without final format)

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}
\usepackage{epsfig}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

\newcommand{\francesc}[1]{\textcolor{blue}{#1}}
\newcommand{\cc}[1]{\textcolor{purple}{#1}}
\newcommand{\francesco}[1]{\textcolor{orange}{#1}}

@@

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the ,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

@@

\begin{abstract}
Spatial Reuse (SR) has recently gained attention \textcolor{orange}{to maximize the performance of} IEEE 802.11 Wireless Local Area Networks (WLANs). Decentralized mechanisms are expected to be key in the development of SR solutions for next-generation WLANs, since many deployments are characterized by being uncoordinated by nature. However, the potential of decentralized mechanisms is limited by the significant lack of knowledge with respect to the overall wireless environment. To shed some light on this subject, we show the main considerations and possibilities of applying online learning to address the SR problem in uncoordinated WLANs. In particular, we provide a solution based on Multi-Armed Bandits (MABs) whereby independent WLANs dynamically adjust their frequency channel, transmit power and sensitivity threshold. To that purpose, we provide two different strategies, which refer to selfish and environment-aware learning. While the former stands for pure individual behavior, the second one \textcolor{orange}{considers} the performance experienced by surrounding networks, thus taking into account the impact of individual actions on the environment. Through these two strategies we delve into practical issues of applying MABs in wireless networks, such as convergence guarantees or adversarial effects. Our simulation results illustrate the potential of the proposed solutions for enabling SR in future WLANs. \textcolor{orange}{We show that substantial improvements on network performance can be achieved regarding throughput and fairness.}
\end{abstract}

@@

%\vfill
%Francesc Wilhelmi (e-mail \href{francisco.wilhelmi@upf.edu}{francisco.wilhelmi@upf.edu}), Sergio Barrachina-Mu\~noz and Boris Bellalta are with the Wireless Networking Research Group, Universitat Pompeu Fabra, 08002 Barcelona, Spain. Cristina Cano is with the Wireless Networks Research Group, Universitat Oberta de Catalunya (UOC), 08860 Castelldefels (Barelona), Spain. Anders~Jonsson and Gergely~Neu are with the Artificial Intelligence and Machine Learning Research Group, Universitat Pompeu Fabra (UPF), 08002 Barcelona, Spain.

\newpage
%% \linenumbers

%%%%%%%%%%%%%%%%%%%%%%%
%%%  I. INTRODUCTION 
%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{section:introduction}
% no \IEEEPARstart	
Wireless communications are rapidly evolving to satisfy the increasingly tighter requirements coming from the explosive growth of wireless devices. To solve that, future Wireless Networks (WNs) are foreseen to cover small areas in high-density scenarios, which evidences the need \textcolor{orange}{for} novel mechanisms to maximize spectral efficiency. In particular, Spatial Reuse (SR) has gained attention in recent years as a potential solution to improve the use of the spectrum. One of the most prominent examples can be found in the IEEE 802.11ax-2019 (11ax) amendment (\citealp{bellalta2016ax}), which defines High-Efficiency (HE) Wireless Local Area Networks (WLANs). \textcolor{orange}{The 11ax amendment aims to maximize spectral efficiency through the SR operation and other} spectrum-efficient techniques like Orthogonal Frequency-Division Multiple Access (OFDMA), and Uplink/Downlink Multi-User Multiple-Input-Multiple-Output (MU-MIMO).

In this paper, we focus on IEEE 802.11 WLANs, which mostly represent uncoordinated deployments (e.g., residential buildings). \textcolor{orange}{These networks are limited in performance because of the scalability issues arising from the current decentralized channel access mechanisms, i.e., Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) (\citealp{ergin2007understanding}).} To enable SR in WLANs, we consider the use of Transmit Power Control (TPC) and Carrier Sense Threshold (CST) adjustment. These mechanisms are of particular concern to the 11ax amendment. \textcolor{orange}{They facilitate and regulate the SR operation by providing a set of procedures and constraints for dynamically setting the transmit power and the sensitivity.} Roughly, the idea of TPC lies in adjusting the transmit power for reducing the interference and/or saving energy. Similarly, CST adjustment seeks to increase the number of parallel transmissions by modifying the sensitivity on a per-device basis. In the context of IEEE 802.11ax, the Dynamic Sensitivity Control (DSC) has been proposed as a potential solution for enabling SR through sensitivity adjustment (\citealp{smith2015dynamic}).

\textcolor{orange}{However, addressing the SR problem in WLANs through TPC and CST adjustment is not trivial for decentralized deployments.} This is mostly motivated by \emph{i)} the spatial interactions between nodes, and \emph{ii)} the adversarial setting unleashed in decentralized wireless networks. For the former, tuning either the transmit power or the sensitivity entails dealing with the spatial dimension. Unlike for frequency and temporal approaches, spatial interference cannot be treated as a binary model. \textcolor{orange}{In the latter case, one can observe when devices transmit or not, thus obtaining full or null performance respectively.} As shown later in Section \ref{section:interactions_wlans}, more complex interactions occur between WLANs tuning their transmit power and CST. As a result, modeling inter-WLANs interactions in next-generation deployments turns out to be extremely complex. 
%For instance, adjusting the CST in a WN brings out the trade-off between the number of parallel transmissions and the collision probability by hidden nodes \cite{afaqui2015evaluation}. 
Moreover, regarding the adversarial setting unleashed by decentralized deployments, strong competition between independent networks may occur.

% OL application
In order to address the SR problem in high-density decentralized networks, we focus on the Multi-Armed Bandit (MAB) framework. \textcolor{orange}{The MAB approach frames the learning-by-interaction problem, and allows to properly approach the exploration-exploitation trade-off in face of uncertainty.} In MABs, a learner (or agent) obtains information from the environment, which reacts according to the actions performed - in the SR problem, an action may refer to a certain configuration of transmit power and CST. By interacting with the environment, a given learner aims to maximize a numerical cumulative reward over time. Unlike classical Reinforcement Learning (RL), the MAB setting does not consider states in general.\footnote{\textcolor{orange}{There is a class of bandits problems that consider states (stateful bandits), which is not considered due to the characteristics of the problem addressed in this work. Essentially, in this class of bandits, states are usually represented by taking strong assumptions that hinder the accuracy of the analysis.}} \textcolor{orange}{A} state may refer to a concrete temporal situation in which the learner is involved. \textcolor{orange}{Therefore, it allows} the latter to construct a policy that determines the behavior to be followed in future situations. Accordingly, learning through states adds an extra layer of complexity and requires that a given agent learns additional contextual information.

The application of MABs into wireless communications problems has recently become very popular (\citealp{chen2010distributed, maghsudi2015channel, maghsudi2015joint}). To model the SR problem through MABs, we consider that WLANs are empowered with an agent that attempts to learn the best-performing action (i.e., a combination of the frequency channel, the transmit power and the sensitivity level). \textcolor{orange}{Its learning operation is based on} the performance achieved in an unknown environment. In this way, MABs operate on the top of CSMA/CA, which operation is influenced by the spatial interactions generated by the taken actions. As a result, we expect WLANs to autonomously find their best configuration in an adversarial setting, given a performance maximization strategy.

% Contributions
The main goal of this paper, then, is to determine the feasibility of applying decentralized learning to improve spectral efficiency in next-generation wireless deployments. In particular, we apply online learning mechanisms to enable SR in dense and uncoordinated WLANs, and show the main derived implications and considerations. We highlight the impact on the aggregate performance and fairness experienced by WLANs, as well as on the guarantees for converging to the optimal solution. The implications of applying online learning to WLANs are studied through the utilization of selfish and environment-aware learning-based strategies. While a selfish strategy is based on the individual performance of a given learner, environment-aware considers \textcolor{orange}{a set of neighboring WLANs that share a reward.} \textcolor{orange}{The SR problem presented in this work is non-convex, therefore prevents to provide any kind of convergence bound. However, our results suggest that the performance of WLANs can be maximized by using learning strategies that are based on probabilistic models of the reward.} To summarize, the main contributions of this work are described next:
%In particular, the SR problem presented here is non-convex, which prevents to provide any kind of convergence bound. Moreover, asymmetries in WNs are very prone to occur due to the random spatial distribution of nodes, and, as further detailed later in Section \ref{section:interactions_wlans}, may have severe implications on the performance of WLANs. 
\begin{itemize}
	\item We showcase the major inter-WLAN dependencies when modifying both the transmit power and the CST, and \textcolor{orange}{how they affect the network performance.}
	\item In order to capture the CSMA/CA operation of IEEE 802.11 WLANs, we use \textcolor{orange}{Continuous Time Markov Networks (CTMNs)} in spatially-distributed scenarios (\citealp{bellalta2017throughput}). We show that CTMNs models are able to capture the existing dependencies between overlapping WLANs. 	
	\item We model the SR problem in WLANs through MABs, where agents implementing Thompson sampling decide the configuration of a given network in terms of frequency channel, transmit power and CST. To the best of our knowledge, this is the first work applying MABs on a CSMA-based network.
	\item We provide insights on the main considerations of using learning in decentralized and adversarial wireless networks. In particular, we showcase the implications of applying selfish and environment-aware learning in dense WLANs, thus emphasizing on the main potentials and pitfalls.
	\item We evaluate the performance of using \textcolor{blue}{self-configuring} agents in dense WLANs, both in specific and random scenarios. The two learning approaches presented in this paper are shown to significantly improve the performance achieved by WLANs in terms of throughput and fairness, with respect to a default - and static - configuration.
\end{itemize}

% Document Structure
The remaining of this document is structured as follows: Section \ref{section:previous_work} refers to the previous related work. In Section \ref{section:interactions_wlans} we first provide details on the throughput model considered in this paper to fit the SR problem. Then, we characterize inter-WLAN interactions when tuning both the transmit power and the sensitivity threshold. \textcolor{orange}{A set of illustrative scenarios is used for that purpose.} Section \ref{section:mabs} formulates the SR problem through MABs and shows the main implications to be considered when applying learning to decentralized wireless networks. Then, Section \ref{section:performance_evaluation} shows the main results of this paper with regard to selfish and environment-aware learning in WLANs. Finally, some remarks are given in Section \ref{section:conclusions}.

@@

To the best of our knowledge, one of the first works to apply RL into a SR-related problem in wireless networks is (\citealp{nie1999q}), in which the authors show a centralized Q-learning mechanism to dynamically select the channel in mobile networks. Other RL-based approaches for channel access can be found in (\citealp{li2009multi, bennis2010q, bennis2011distributed, sallent2015learning, rupasinghe2015reinforcement}), covering cognitive radio, self-organizing cellular networks and coexistence problems. Despite Q-learning (or other Markovian-based methods) has been shown to properly fit to channel allocation problems, few applications have been provided to the SR problem. Note, as well, that dealing with the frequency domain allows to naturally define states,\footnote{An state describes a particular situation of a given agent at a specific time.} which can be done according to the availability of channels (typically modeled through Bernoulli distributions). Therefore, an agent may observe the environment and define \textcolor{orange}{an} accurate model where the state is defined by the set of channels that are available/occupied. Note that the contextual information provided to the learner is important for learning efficiently, since the agent is able to \textcolor{orange}{react} to different situations. With a \textcolor{orange}{proper definition of states}, a higher degree of control is conferred to the agent. Therefore, provided that the \textcolor{orange}{model of the states} is accurate enough, the learning procedure carried out by a given learner can result into \textcolor{orange}{better performance than that of} a stateless setting.

However, modeling states for the decentralized SR problem \textcolor{orange}{entails added complexity}, thus hindering the learning procedure followed by a given agent. In the particular case of IEEE 802.11 WLANs, spatial interactions among nodes lead to complex scenarios\textcolor{orange}{. The} performance achieved by a given WLAN depends on the additive interference \textcolor{orange}{coming} from an unknown environment. Therefore, learning accurate enough states for the SR problem turns out to be challenging. Note that, if states do not reflect the actual situation of a given agent at a given moment, the learnings that can be generated become strongly limited, and can even be meaningless. To cope with the difficulties on modeling states for the decentralized SR problem, we focus on multi-player MABs (MP-MABs), which frame resource allocation problems where several agents compete against each other. MP-MABs have been recently broadly applied for opportunistic spectrum access in cognitive radio (\citealp{liu2010distributed, anandkumar2011distributed, rosenski2016multi, maghsudi2015joint, maghsudi2015channel}).

Firstly, in (\citealp{liu2010distributed}), the authors provide a decentralized policy with logarithmic regret order, which is based on a time-division fair sharing of the best arms. However, such a policy requires coordination among agents and to know the exact number of adversarial nodes, which in addition must be constant and known in advance. Both requirements entail dedicated communication between nodes, which turns out to be unfeasible for decentralized problems such as the one presented in this paper. Another important contribution regarding multi-player learning for the opportunistic spectrum access problem is provided in (\citealp{anandkumar2011distributed}), where the authors provide a distributed learning algorithm that showcases order-optimal regret. However, the total number of secondary users is known by the system, which may not be feasible in real scenarios where no communication between nodes exists. In contrast, in this work we consider selfish and environment-aware learning approaches, none of which require explicit communication between independent learners. Furthermore, a less strict method \textcolor{orange}{is also provided in (\citealp{anandkumar2011distributed})}, in which the number of secondary users (which is fixed) is estimated, so that nearly order-optimal regret is achieved. In both algorithms, it is assumed that all the users use the same policy. Regarding the work in (\citealp{rosenski2016multi}), sublinear regret is achieved if all players implement the proposed algorithm. Some interesting thoughts are provided regarding varying environments, which, to the best of our knowledge, have been barely considered in the previous literature. %(note, as well, that varying scenarios are neither considered in this work).
 For instance, the authors emphasize that, in the dynamic setting, the frequency at which players enter and leave the scenario must be limited in order to provide a sublinear regret. Unlike the SR presented in this paper, the work in (\citealp{rosenski2016multi}) assumes that there exists an optimal solution whereby no collisions occur for any player. The concept of collision is inspired in the ALOHA channel access mechanism, and occurs if two or more players choose the same arm (or channel). %Moreover, even if communication between nodes is not considered, synchronized clocks are required in order to regulate the time period in which arrivals and departures of learners occur.

When it comes to the SR problem \textcolor{orange}{by means of TPC and/or sensitivity adjustment}, we find MP-MABs application for joint channel selection and power control in (\citealp{maghsudi2015joint, maghsudi2015channel}). The work in (\citealp{maghsudi2015joint}) proposes a strategy based on the Signal-to-Interference-plus-Noise Ratio (SINR) to determine the channel and the transmit power to be used in cognitive radio networks. The authors prove that the MP-MAB game converges to a correlated equilibrium, which in addition maximizes the aggregate utility, if the two following assumptions hold: \emph{i)} the problem is relaxed, so that the reward granted to a given agent depends only on the Signal-to-Noise Ratio (SNR), regardless of the overlapping interference, \emph{ii)} user-specific penalties are provided to each agent. \textcolor{orange}{In contrast with} the work in (\citealp{maghsudi2015joint}), here we aim to understand the potentials and limitations of applying MABs in a CSMA/CA-based setting, in which the previous assumptions do not hold. First of all, instead of relaxing the problem to only consider the SNR, we model the interactions at the MAC level. Secondly, defining user-specific penalties would require the use of either a centralized system or message exchanging.

Finally, in (\citealp{maghsudi2015channel}), the authors introduce the concept of calibrated forecaster, i.e., a predictor of the \textcolor{orange}{actions of the adversaries} that improves with collected knowledge. By using such a predictor, if every learner is able to predict and respond to the others' actions, then the game converges to a correlated equilibrium. In other words, if a node can predict which channel will the adversary pick (and vice-versa), then it can select the other channel and experience the maximum performance. In contrast, \textcolor{orange}{in the SR problem we tackle here}, developing an accurate forecaster in a decentralized way may be an extremely complex task. Refer to the non-linear relationships that occur in the spatial domain, which are hard to model, and thus to predict. Moreover, density and messy deployments may remove the existence of an equilibrium, hence invalidating the assumptions.
%and the presence of legacy nodes (i.e., nodes not using the forecaster) may harm the joint operation carried out by learners.

In summary, a lot of effort has been recently made to enable the evolution of wireless networks towards self-adjusting systems. In particular, the application of RL has been extensively studied for channel access problems. However, these type of methods do not properly suit to spatially-distributed problems such as the SR one. As a result, other stateless techniques, such as MABs, have been targeted, and have shown to effectively improve wireless networks performance, even in adversarial environments. Nevertheless, these mechanisms require that \textcolor{orange}{strong assumptions about the system model hold}. Moreover, spatial interactions between WLANs \textcolor{orange}{have not yet been considered}.

%%%%%%%%%%%%%%%%%%
%%% III. SYSTEM MODEL 
%%%%%%%%%%%%%%%%%%
\section{Interactions between WLANs when Spatial Reuse is Enabled}
\label{section:interactions_wlans}    
In this Section, for completeness, we first briefly introduce the CSMA/CA operation carried out by Wi-Fi networks for accessing the channel, as well as the CSMA/CA throughput model considered along this paper. Therewith, we aim to identify the main inter-WLAN interactions when modifying both the transmit power and the \textcolor{orange}{CST}. Understanding these interactions is key to motivate the usage of MABs to the decentralized SR problem. As shown in Section \ref{section:previous_work}, some of the previous work addressed similar problems and provided mechanisms that were proven to converge to an equilibrium. However, the novelty of this paper lies in the analysis of learning techniques in CSMA/CA-based networks. \textcolor{orange}{Unlike previous} work, where the reward (i.e., the throughput) is mostly given by a linear function that only depends on the signal strength and the interference, here we deal with more complex interaction between networks. In Wi-Fi, due to the decentralized nature of CSMA/CA, an optimal solution in terms of SR is harder \textcolor{orange}{to derive} than in cellular-based networks. In addition, there is a trade-off between performance maximization and fairness, \textcolor{orange}{which is not trivial to compute in a decentralized setting}.

>>>>

\usepackage{epsfig}
\usepackage{amssymb}

@@

@@

\begin{abstract}
Spatial Reuse (SR) has recently gained attention to maximize the performance of IEEE 802.11 Wireless Local Area Networks (WLANs). Decentralized mechanisms are expected to be key in the development of SR solutions for next-generation WLANs, since many deployments are characterized by being uncoordinated by nature. However, the potential of decentralized mechanisms is limited by the significant lack of knowledge with respect to the overall wireless environment. To shed some light on this subject, we show the main considerations and possibilities of applying online learning to address the SR problem in uncoordinated WLANs. In particular, we provide a solution based on Multi-Armed Bandits (MABs) whereby independent WLANs dynamically adjust their frequency channel, transmit power and sensitivity threshold. To that purpose, we provide two different strategies, which refer to selfish and environment-aware learning. While the former stands for pure individual behavior, the second one considers the performance experienced by surrounding networks, thus taking into account the impact of individual actions on the environment. Through these two strategies we delve into practical issues of applying MABs in wireless networks, such as convergence guarantees or adversarial effects. Our simulation results illustrate the potential of the proposed solutions for enabling SR in future WLANs. We show that substantial improvements on network performance can be achieved regarding throughput and fairness.
\end{abstract}

@@

\newpage

%%%%%%%%%%%%%%%%%%%%%%%
%%%  I. INTRODUCTION 
%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{section:introduction}
% no \IEEEPARstart	
Wireless communications are rapidly evolving to satisfy the increasingly tighter requirements coming from the explosive growth of wireless devices. To solve that, future Wireless Networks (WNs) are foreseen to cover small areas in high-density scenarios, which evidences the need for novel mechanisms to maximize spectral efficiency. In particular, Spatial Reuse (SR) has gained attention in recent years as a potential solution to improve the use of the spectrum. One of the most prominent examples can be found in the IEEE 802.11ax-2019 (11ax) amendment (\citealp{bellalta2016ax}), which defines High-Efficiency (HE) Wireless Local Area Networks (WLANs). The 11ax amendment aims to maximize spectral efficiency through the SR operation and other spectrum-efficient techniques like Orthogonal Frequency-Division Multiple Access (OFDMA), and Uplink/Downlink Multi-User Multiple-Input-Multiple-Output (MU-MIMO).

In this paper, we focus on IEEE 802.11 WLANs, which mostly represent uncoordinated deployments (e.g., residential buildings). These networks are limited in performance because of the scalability issues arising from the current decentralized channel access mechanisms, i.e., Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA) (\citealp{ergin2007understanding}). To enable SR in WLANs, we consider the use of Transmit Power Control (TPC) and Carrier Sense Threshold (CST) adjustment. These mechanisms are of particular concern to the 11ax amendment. They facilitate and regulate the SR operation by providing a set of procedures and constraints for dynamically setting the transmit power and the sensitivity. Roughly, the idea of TPC lies in adjusting the transmit power for reducing the interference and/or saving energy. Similarly, CST adjustment seeks to increase the number of parallel transmissions by modifying the sensitivity on a per-device basis. In the context of IEEE 802.11ax, the Dynamic Sensitivity Control (DSC) has been proposed as a potential solution for enabling SR through sensitivity adjustment (\citealp{smith2015dynamic}).

However, addressing the SR problem in WLANs through TPC and CST adjustment is not trivial for decentralized deployments. This is mostly motivated by \emph{i)} the spatial interactions between nodes, and \emph{ii)} the adversarial setting unleashed in decentralized wireless networks. For the former, tuning either the transmit power or the sensitivity entails dealing with the spatial dimension. Unlike for frequency and temporal approaches, spatial interference cannot be treated as a binary model. In the latter case, one can observe when devices transmit or not, thus obtaining full or null performance respectively. As shown later in Section \ref{section:interactions_wlans}, more complex interactions occur between WLANs tuning their transmit power and CST. As a result, modeling inter-WLANs interactions in next-generation deployments turns out to be extremely complex. Moreover, regarding the adversarial setting unleashed by decentralized deployments, strong competition between independent networks may occur.

% OL application
In order to address the SR problem in high-density decentralized networks, we focus on the Multi-Armed Bandit (MAB) framework. The MAB approach frames the learning-by-interaction problem, and allows to properly approach the exploration-exploitation trade-off in face of uncertainty. In MABs, a learner (or agent) obtains information from the environment, which reacts according to the actions performed - in the SR problem, an action may refer to a certain configuration of transmit power and CST. By interacting with the environment, a given learner aims to maximize a numerical cumulative reward over time. Unlike classical Reinforcement Learning (RL), the MAB setting does not consider states in general.\footnote{There is a class of bandits problems that consider states (stateful bandits), which is not considered due to the characteristics of the problem addressed in this work. Essentially, in this class of bandits, states are usually represented by taking strong assumptions that hinder the accuracy of the analysis.} A state may refer to a concrete temporal situation in which the learner is involved. Therefore, it allows the latter to construct a policy that determines the behavior to be followed in future situations. Accordingly, learning through states adds an extra layer of complexity and requires that a given agent learns additional contextual information.

The application of MABs into wireless communications problems has recently become very popular (\citealp{chen2010distributed, maghsudi2015channel, maghsudi2015joint}). To model the SR problem through MABs, we consider that WLANs are empowered with an agent that attempts to learn the best-performing action (i.e., a combination of the frequency channel, the transmit power and the sensitivity level). Its learning operation is based on the performance achieved in an unknown environment. In this way, MABs operate on the top of CSMA/CA, which operation is influenced by the spatial interactions generated by the taken actions. As a result, we expect WLANs to autonomously find their best configuration in an adversarial setting, given a performance maximization strategy.

% Contributions
The main goal of this paper, then, is to determine the feasibility of applying decentralized learning to improve spectral efficiency in next-generation wireless deployments. In particular, we apply online learning mechanisms to enable SR in dense and uncoordinated WLANs, and show the main derived implications and considerations. We highlight the impact on the aggregate performance and fairness experienced by WLANs, as well as on the guarantees for converging to the optimal solution. The implications of applying online learning to WLANs are studied through the utilization of selfish and environment-aware learning-based strategies. While a selfish strategy is based on the individual performance of a given learner, environment-aware considers a set of neighboring WLANs that share a reward. The SR problem presented in this work is non-convex, therefore prevents to provide any kind of convergence bound. However, our results suggest that the performance of WLANs can be maximized by using learning strategies that are based on probabilistic models of the reward. To summarize, the main contributions of this work are described next:
%In particular, the SR problem presented here is non-convex, which prevents to provide any kind of convergence bound. Moreover, asymmetries in WNs are very prone to occur due to the random spatial distribution of nodes, and, as further detailed later in Section \ref{section:interactions_wlans}, may have severe implications on the performance of WLANs. 
\begin{itemize}
	\item We showcase the major inter-WLAN dependencies when modifying both the transmit power and the CST, and how they affect the network performance.
	\item In order to capture the CSMA/CA operation of IEEE 802.11 WLANs, we use Continuous Time Markov Networks (CTMNs) in spatially-distributed scenarios (\citealp{bellalta2017throughput}). We show that CTMNs models are able to capture the existing dependencies between overlapping WLANs. 	
	\item We model the SR problem in WLANs through MABs, where agents implementing Thompson sampling decide the configuration of a given network in terms of frequency channel, transmit power and CST. To the best of our knowledge, this is the first work applying MABs on a CSMA-based network.
	\item We provide insights on the main considerations of using learning in decentralized and adversarial wireless networks. In particular, we showcase the implications of applying selfish and environment-aware learning in dense WLANs, thus emphasizing on the main potentials and pitfalls.
	\item We evaluate the performance of using self-configuring agents in dense WLANs, both in specific and random scenarios. The two learning approaches presented in this paper are shown to significantly improve the performance achieved by WLANs in terms of throughput and fairness, with respect to a default - and static - configuration.
\end{itemize}

% Document Structure
The remaining of this document is structured as follows: Section \ref{section:previous_work} refers to the previous related work. In Section \ref{section:interactions_wlans} we first provide details on the throughput model considered in this paper to fit the SR problem. Then, we characterize inter-WLAN interactions when tuning both the transmit power and the sensitivity threshold. A set of illustrative scenarios is used for that purpose. Section \ref{section:mabs} formulates the SR problem through MABs and shows the main implications to be considered when applying learning to decentralized wireless networks. Then, Section \ref{section:performance_evaluation} shows the main results of this paper with regard to selfish and environment-aware learning in WLANs. Finally, some remarks are given in Section \ref{section:conclusions}.

@@

To the best of our knowledge, one of the first works to apply RL into a SR-related problem in wireless networks is (\citealp{nie1999q}), in which the authors show a centralized Q-learning mechanism to dynamically select the channel in mobile networks. Other RL-based approaches for channel access can be found in (\citealp{li2009multi, bennis2010q, bennis2011distributed, sallent2015learning, rupasinghe2015reinforcement}), covering cognitive radio, self-organizing cellular networks and coexistence problems. Despite Q-learning (or other Markovian-based methods) has been shown to properly fit to channel allocation problems, few applications have been provided to the SR problem. Note, as well, that dealing with the frequency domain allows to naturally define states,\footnote{An state describes a particular situation of a given agent at a specific time.} which can be done according to the availability of channels (typically modeled through Bernoulli distributions). Therefore, an agent may observe the environment and define an accurate model where the state is defined by the set of channels that are available/occupied. Note that the contextual information provided to the learner is important for learning efficiently, since the agent is able to react to different situations. With a proper definition of states, a higher degree of control is conferred to the agent. Therefore, provided that the model of the states is accurate enough, the learning procedure carried out by a given learner can result into better performance than that of a stateless setting.

However, modeling states for the decentralized SR problem entails added complexity, thus hindering the learning procedure followed by a given agent. In the particular case of IEEE 802.11 WLANs, spatial interactions among nodes lead to complex scenarios. The performance achieved by a given WLAN depends on the additive interference coming from an unknown environment. Therefore, learning accurate enough states for the SR problem turns out to be challenging. Note that, if states do not reflect the actual situation of a given agent at a given moment, the learnings that can be generated become strongly limited, and can even be meaningless. To cope with the difficulties on modeling states for the decentralized SR problem, we focus on multi-player MABs (MP-MABs), which frame resource allocation problems where several agents compete against each other. MP-MABs have been recently broadly applied for opportunistic spectrum access in cognitive radio (\citealp{liu2010distributed, anandkumar2011distributed, rosenski2016multi, maghsudi2015joint, maghsudi2015channel}).

Firstly, in (\citealp{liu2010distributed}), the authors provide a decentralized policy with logarithmic regret order, which is based on a time-division fair sharing of the best arms. However, such a policy requires coordination among agents and to know the exact number of adversarial nodes, which in addition must be constant and known in advance. Both requirements entail dedicated communication between nodes, which turns out to be unfeasible for decentralized problems such as the one presented in this paper. Another important contribution regarding multi-player learning for the opportunistic spectrum access problem is provided in (\citealp{anandkumar2011distributed}), where the authors provide a distributed learning algorithm that showcases order-optimal regret. However, the total number of secondary users is known by the system, which may not be feasible in real scenarios where no communication between nodes exists. In contrast, in this work we consider selfish and environment-aware learning approaches, none of which require explicit communication between independent learners. Furthermore, a less strict method is also provided in (\citealp{anandkumar2011distributed}), in which the number of secondary users (which is fixed) is estimated, so that nearly order-optimal regret is achieved. In both algorithms, it is assumed that all the users use the same policy. Regarding the work in (\citealp{rosenski2016multi}), sublinear regret is achieved if all players implement the proposed algorithm. Some interesting thoughts are provided regarding varying environments, which, to the best of our knowledge, have been barely considered in the previous literature. For instance, the authors emphasize that, in the dynamic setting, the frequency at which players enter and leave the scenario must be limited in order to provide a sublinear regret. Unlike the SR presented in this paper, the work in (\citealp{rosenski2016multi}) assumes that there exists an optimal solution whereby no collisions occur for any player. The concept of collision is inspired in the ALOHA channel access mechanism, and occurs if two or more players choose the same arm (or channel).

When it comes to the SR problem by means of TPC and/or sensitivity adjustment, we find MP-MABs application for joint channel selection and power control in (\citealp{maghsudi2015joint, maghsudi2015channel}). The work in (\citealp{maghsudi2015joint}) proposes a strategy based on the Signal-to-Interference-plus-Noise Ratio (SINR) to determine the channel and the transmit power to be used in cognitive radio networks. The authors prove that the MP-MAB game converges to a correlated equilibrium, which in addition maximizes the aggregate utility, if the two following assumptions hold: \emph{i)} the problem is relaxed, so that the reward granted to a given agent depends only on the Signal-to-Noise Ratio (SNR), regardless of the overlapping interference, \emph{ii)} user-specific penalties are provided to each agent. In contrast with the work in (\citealp{maghsudi2015joint}), here we aim to understand the potentials and limitations of applying MABs in a CSMA/CA-based setting, in which the previous assumptions do not hold. First of all, instead of relaxing the problem to only consider the SNR, we model the interactions at the MAC level. Secondly, defining user-specific penalties would require the use of either a centralized system or message exchanging.

Finally, in (\citealp{maghsudi2015channel}), the authors introduce the concept of calibrated forecaster, i.e., a predictor of the actions of the adversaries that improves with collected knowledge. By using such a predictor, if every learner is able to predict and respond to the others' actions, then the game converges to a correlated equilibrium. In other words, if a node can predict which channel will the adversary pick (and vice-versa), then it can select the other channel and experience the maximum performance. In contrast, in the SR problem we tackle here, developing an accurate forecaster in a decentralized way may be an extremely complex task. Refer to the non-linear relationships that occur in the spatial domain, which are hard to model, and thus to predict. Moreover, density and messy deployments may remove the existence of an equilibrium, hence invalidating the assumptions.
%and the presence of legacy nodes (i.e., nodes not using the forecaster) may harm the joint operation carried out by learners.

In summary, a lot of effort has been recently made to enable the evolution of wireless networks towards self-adjusting systems. In particular, the application of RL has been extensively studied for channel access problems. However, these type of methods do not properly suit to spatially-distributed problems such as the SR one. As a result, other stateless techniques, such as MABs, have been targeted, and have shown to effectively improve wireless networks performance, even in adversarial environments. Nevertheless, these mechanisms require that strong assumptions about the system model hold. Moreover, spatial interactions between WLANs have not yet been considered.

%%%%%%%%%%%%%%%%%%
%%% III. SYSTEM MODEL 
%%%%%%%%%%%%%%%%%%
\section{Interactions between WLANs when Spatial Reuse is Enabled}
\label{section:interactions_wlans}    
In this Section, for completeness, we first briefly introduce the CSMA/CA operation carried out by Wi-Fi networks for accessing the channel, as well as the CSMA/CA throughput model considered along this paper. Therewith, we aim to identify the main inter-WLAN interactions when modifying both the transmit power and the CST. Understanding these interactions is key to motivate the usage of MABs to the decentralized SR problem. As shown in Section \ref{section:previous_work}, some of the previous work addressed similar problems and provided mechanisms that were proven to converge to an equilibrium. However, the novelty of this paper lies in the analysis of learning techniques in CSMA/CA-based networks. Unlike previous work, where the reward (i.e., the throughput) is mostly given by a linear function that only depends on the signal strength and the interference, here we deal with more complex interaction between networks. In Wi-Fi, due to the decentralized nature of CSMA/CA, an optimal solution in terms of SR is harder to derive than in cellular-based networks. In addition, there is a trade-off between performance maximization and fairness, which is not trivial to compute in a decentralized setting.
